---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---


## Loading and preprocessing the data
The data is loaded from the csv file in character format. The individual fields are converted into appropriate data classes
```{r loadandpreprocess}
if(file.exists("activity.csv") != TRUE)
{
    unzip("activity.zip")
}

actData <- read.csv("activity.csv", colClasses="character")
actData$steps <- as.numeric(actData$steps)
actData$date <- as.Date(actData$date, "%Y-%m-%d")
actData$interval <- as.numeric(actData$interval)
intervalLevels <- as.numeric(levels(factor(actData$interval)))
```


## What is mean total number of steps taken per day?
The total steps taken per day is calculated and histogram of the same are plotted below
```{r stepSum}
naRmSum <- function(x) sum(x, na.rm = T)
stepSum <- tapply(actData$steps, factor(actData$date), naRmSum)
hist(stepSum, main = "Histogram of Total Daily Steps", xlab = 
         "Number of Steps", ylab = "Frequency")
stepSumMean <- mean(stepSum, na.rm = T)
stepSumMedian <- median(stepSum, na.rm = T)
```
The mean and median values of the total number of steps per day are `r stepSumMean` and `r stepSumMedian` respectively.


## What is the average daily activity pattern?
The averages of the number of steps is calculated per interval and a plot of the same is below
```{r intevalStepPattern}
naRmMean <- function(x) mean(x, na.rm = T)
intervalMean <- tapply(actData$steps, factor(actData$interval), naRmMean)
plot(intervalMean~levels(factor(actData$interval)), type="l", main = 
         "Daily Activity Pattern", xlab = "Interval", ylab = 
         "Average Number Of Steps")
maxIntervalIndex <- which(intervalMean == max(intervalMean))
maxInterval <- intervalLevels[maxIntervalIndex]
```
The interval with the maximum number of steps on average is `r maxInterval`


## Imputing missing value
The number of missing values in the measurements is calculated. The average steps for each time interval is calculated and is used to replace the missing values. This will compensate for the lack of data more accurately.
```{r removeMissing}
numMissingVal <- length(which(is.na(actData$steps)))
actDataNew <- actData
for(i in 1:length(actData$steps))
{
    if(is.na(actData$steps[i]))
    {
        intervalVal <- actData$interval[i] 
        actDataNew$steps[i] <- intervalMean[[which(intervalLevels == intervalVal)]]
    }
}
stepSumNew <- tapply(actDataNew$steps, factor(actDataNew$date), naRmSum)
hist(stepSumNew, main = "Histogram of Total Daily Steps", xlab = 
         "Number of Steps", ylab = "Frequency")
stepSumMeanNew <- mean(stepSumNew, na.rm = T)
stepSumMedianNew <- median(stepSumNew, na.rm = T)

```
The new mean and median values of the total number of steps per day are `r stepSumMeanNew` and `r stepSumMedianNew` respectively.

## Are there differences in activity patterns between weekdays and weekends?
The activity patterns for weekends and weekdays are calculated and plotted below. It is observed that weekends have a higher activity as indicated by higher frequency of higher ranges of steps taken
```{r weekday}
actDataNew$day <- weekdays(as.Date(actDataNew$date))
for(i in 1:length(actDataNew$date))
{
    if((actDataNew$day[i] == "Saturday") | (actDataNew$day[i] == "Sunday"))
    {
        actDataNew$wdWe[i] <- "Weekend"
    }
    else
    {
        actDataNew$wdWe[i] <- "Weekday"
    }
}
splitActData <- split(actDataNew, actDataNew$wdWe)

intervalMeanWd <- tapply(splitActData[[1]]$steps, 
                         factor(splitActData[[1]]$interval), naRmMean)
intervalMeanWe <- tapply(splitActData[[2]]$steps, 
                         factor(splitActData[[2]]$interval), naRmMean)

plot(intervalMeanWd~levels(factor(actDataNew$interval)), type= "l", main = 
         "Weekday Activity Pattern", xlab = "Interval", ylab = 
         "Average Number Of Steps")
plot(intervalMeanWe~levels(factor(actDataNew$interval)), type = "l", main =
         "Weekend Activity Pattern", xlab = "Interval", ylab = 
         "Average Number Of Steps")
```